Public transport and school location impacts on educational inequalities: insights from São Paulo

### 2017-04-02 paper_v2.tex
Justify emphasis on PT

The SPMR presents an excellent opportunity to measure the extent of
educational and public transport provision, and to study the impact of
public policies on different socio-economic groups. Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015. However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016. The change in policy was estimated to affect more than 300,000 students, many whom would be placed in schools far from their homes. As a response, in the following months students occupied over 200 schools. Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (Ortellado 2015).
It was this policy context that informed the paper's focus on public transport, rather than the topic of active travel to school, which has gained recent prominence in Australia (e.g. Veitch et al., 2017) the USA and Europe

Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas. In this dynamic and highly political context, new quantitative evidence can help shed light on the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing. Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the
provision of schools, travel to school opportunities, or both, are
particularly restrictive.

@@

@@

>>>>

The SPMR presents an excellent opportunity to measure the extent of
educational and public transport provision, and to study the impact of
public policies on different socio-economic groups. Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015. However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016. The change in policy was estimated to affect more than 300,000 students, many whom would be placed in schools far from their homes. As a response, in the following months students occupied over 200 schools. Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (Ortellado 2015).

Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas. In this dynamic and highly political context, new quantitative evidence can help shed light on the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing. Moreover, the results should help design school location and public transport policies that are more inclusive and equitable by highlighting areas where
school location and public transport opportunities conspire to exacerbate educational inequalities.

The recent experience of São Paulo and other cities informed the paper's public transport emphasis. 
This could be seen as a contrast to the topic of active travel to school, which has gained recent prominence in Australia (e.g. Veitch et al., 2017) the USA (Lee et al., 2017) and Europe (e.g. Macdonald et al., 2016).
However, we see great potential synergies between provision of public and active transport opportunities and, by focussing on public transport, do not wish to endorse an `either or' approach to adequate school travel opportunities.
The focus on public transport is especially appropriate in a Latin American context where parents' perception of road danger may (rightly) be greater and where parents cannot be expected automatically to be able to have the resources to buy and maintain a bike for children to cycle to school (public provision of safe walking and cycling infrastructure and bicycles are areas of research worthy of attention beyond the scope of this paper).

@@

Lee, Richard J., Ipek N. Sener, and S. Nathan Jones. 2017. ``Understanding the Role of Equity in Active Transportation Planning in the United States.'' \emph{Transport Reviews} 37 (2): 211–26.

@@

Macdonald, Laura, Paul McCrorie, Natalie Nicholls, and Anne Ellaway. 2016. ``Walkability around Primary Schools and Area Deprivation across Scotland.'' \emph{BMC Public Health} 16: 328.
